---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/hypertidy/geodist.svg)](https://travis-ci.org/hypertidy/geodist) 
[![Project Status: Concept - Minimal or no implementation has been done yet.](http://www.repostatus.org/badges/0.1.0/concept.svg)](http://www.repostatus.org/#concept)
[![codecov](https://codecov.io/gh/hypertidy/geodist/branch/master/graph/badge.svg)](https://codecov.io/gh/hypertidy/geodist)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/geodist)](http://cran.r-project.org/web/packages/geodist) 
![downloads](http://cranlogs.r-pkg.org/badges/grand-total/geodist)

# geodist

An ultra-lightweight, zero-dependency package for calculation of geodesic
distances.

## Installation

You can install geodist from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("hypertidy/geodist")
```

## Usage

`geodist` contains only one eponymous function which accepts only one or two
primary arguments, each of which must be some kind of rectangular object with
unambiguously labelled longitude and latitude columns (that is, some variant of
`lon`/`lat`, or `x`/`y`). The only other argument is `measure`, specifying one
of Haversine or Vincenty great-circle distances.

```{r, echo = FALSE}
devtools::load_all (".", export_all = FALSE)
```
```{r, eval=FALSE}
library(geodist)
```
```{r}
# current verison
packageVersion("geodist")
```



```{r}
n <- 1e1
x <- tibble::tibble (x = -180 + 360 * runif (n),
                     y = -90 + 180 * runif (n))
dim (geodist (x))
y <- tibble::tibble (x = -180 + 360 * runif (2 * n),
                     y = -90 + 180 * runif (2 * n))
dim (geodist (x, y))
```

### Comparison

Geodesic distance calculation is available in the [`sf`
package](https://cran.r-project.org/package=sf), requiring only conversion of
sets of numeric lon-lat points to `sf` form with the following code:
```{r x_to_sf}
require (magrittr)
x_to_sf <- function (x)
{
    sapply (seq (nrow (x)), function (i)
            sf::st_point (x [i, ]) %>%
                sf::st_sfc ()) %>%
    sf::st_sfc ()
}
```
Note that the relevant function, `sf::st_distance()` calculates the Vicenty
distance via `liblwgeom`, so `measure = "vincenty"` is used in the following
comparison.

```{r benchmark}
n <- 1e3
x <- cbind (-180 + 360 * runif (n), -90 + 180 * runif (n))
colnames (x) <- c ("x", "y")
xsf <- x_to_sf (x)
sf_dist <- function (x) sf::st_distance (x, x)
rbenchmark::benchmark (replications = 10, order = "test",
                      sf_dist (xsf),
                      geodist (x, measure = "vincenty")) [, 1:4]
```

The `geodist` package also implements the [mapbox cheap ruler
algorithm](https://github.com/mapbox/cheap-ruler-cpp) (described in this [blog
post](https://blog.mapbox.com/fast-geodesic-approximations-with-cheap-ruler-106f229ad016),
providing approximate yet very fast distance calculations within small
(typically intra-urban) areas. Speed advantages are demonstrated in the
following code:
```{r benchmark2}
n <- 1e3
dx <- dy <- 0.01
x <- cbind (-100 + dx * runif (n), 20 + dy * runif (n))
y <- cbind (-100 + dx * runif (2 * n), 20 + dy * runif (2 * n))
colnames (x) <- colnames (y) <- c ("x", "y")
rbenchmark::benchmark (replications = 10, order = "test",
                       d1 <- geodist (x, measure = "cheap"),
                       d2 <- geodist (x, measure = "haversine"),
                       d3 <- geodist (x, measure = "vincenty")) [, 1:4]
```


### Test Results

```{r load-testthat, message = FALSE}
require (devtools)
require (testthat)
```

```{r tests, eval = TRUE}
date()
devtools::test("tests/")
```
